\textit{MSF: Modulated Sub-graph Finder}

### 2019-03-11 MSF_Paper_FA.tex
last revision fabian

		To overcome this limitation, we developed a novel approach to identify
		concertedly modulated sub-graphs in the global cell signaling network,
		based on the DGEA results of all genes tested. Thereby, expression patterns
		of genes are integrated according to the topology of their interactions and
		allow potentially to read the flow of information and identify the effectors. The described software, named \texttt{Modulated
			Sub-graph Finder} (\texttt{MSF}) is freely available at
		\url{https://github.com/Modulated-Subgraph-Finder/MSF}.

@@

	Existing pathway-based analysis approaches use different research
	designs, which can be categorized into ORA (Over-representation
	analysis), FCS (Functional class scoring) and pathway topology based
	methods. All of which aim to find a subset of genes, e.g.,
	significantly differentially expressed genes, genes associated with a
	certain pathway more often than expected given the total set of
	examined genes, e.g.~the whole genome background. ORA is
	the first and the most basic method of pathway analysis~\cite{Campos}. It uses a
	DEG list with user defined cut-off for the log-fold change and
	\textit{p}-value (most commonly using absolute log-fold change $\geq$
	2 and \textit{p}-value $\leq$ 0.05). Subsequently, sets of genes
	associated with annotated pathways are tested for being
	over-represented in the set of DEGs. To this end, hyper-geometric
	distribution, chi-square tests, binomial probability or the Fisher’s
	exact test are used. Thereby the information of the topology of genes
	in the pathways is neglected~\cite{Bayer}. Furthermore, ORA assumes
	that the biological pathways are independent of each other and ignores
	the fact that they cross-talk and overlap~\cite{Khatri2012,Campos}.

@@

	From another perspective, network based approaches do not categorize sets of genes into functional pathways, but they consider all interactions to be equal. Thereby, they avoid distinguishing arbitrarily between interactions within a pathway and interactions between pathways (i.e., cross-talk). With this they aim to identify sub-networks that show modulation between two conditions or upon a stimuli~\cite{mitra_carvunis_ramesh_ideker_2013}. To find these active modules heuristics solutions like simulated annealing (SA), greedy methods, genetic algorithms (GA), network propagation and co-clustering methods are used~\cite{mitra_carvunis_ramesh_ideker_2013}. jActiveModules has been the first of this kind using simulated annealing to find the modulated sub-networks~\cite{jActiveModules}. The benefit of omitting pathways is bought by reduced interpretability of the results due to the lack of functional labels on the networks.

@@

	As input, information on functional links between genes provided by
	e.g.~KEGG or Reactome and information on the differential expression status
	of single genes resulting from a DGEA, are required. As a result the
	analysis returns sub-graphs and their joint confidence scores, reflecting
	how the perturbation is migrated through the network. Furthermore, the
	entry points of perturbation in the networks and overlap with conventional
	pathway categories are returned. These entry points are given impact scores showing the degree of perturbations introduced by them. The reliability of the entry points is checked using a t-test on the \textit{p}-values by investigating the significant difference between the downstream and upstream genes of each entry point. The output is prepared in a directed
	adjacency file, convenient for visualization, e.g., with
	StringApp~\cite{StringApp}, available as a Cytoscape plug-in~\cite{Cyto}.

	The proposed algorithm is named
	\texttt{Modulated Sub-graph Finder} (abbreviated
	\texttt{MSF}). \texttt{MSF} can help transform the information obtained
	from DGEA into comprehensible knowledge of signal transduction of genes and
	thereby being a valuable complement to existing pathway based
	methods. \texttt{MSF} is freely accessible on git hub under the terms of the
	Creative Commons Attribution 4.0 International License.

@@

	\texttt{MSF} was implemented as a java program. It is developed as a novel heuristic approach to find concertedly
	modulated sub-graphs in networks of biological interactions. \texttt{MSF}
	does not use predefined gene sets grouped into functional units, but rather
	relies purely on the network of interacting genes. The input network
	consists of nodes corresponding to genes and edges representing
	interactions. Furthermore it utilizes comprehensive results from a
	differential gene expression analysis to discover the sub-graphs, or
	modules, which are as a whole modulated.

@@

	Lambda $\lambda$ be the weights for each gene, currently all genes have equal weight i.e. 1. The combined \textit{p}-value  $t(\rho)$ of a sub-graph will express the
	significance of all genes in the sub-graph being modulated
	together. Thereby, the information from the different genes are used as,
	although not independent, replicated measurement of the behavior of the
	whole sub-graph. This potentially increases the power to detect also
	significant sub-graphs consisting of genes which are not significant on
	there own.

@@

	In a last post processing step \texttt{MSF} identifies the
        trigger points of the modulated sub-graphs. These trigger
        genes are the sources of the sub-graphs with only outgoing
        edges. These genes can be interpreted as the possible entry
        points of perturbation from where the stimulus causes
        downstream effects. Each individual source is given an impact
        score, expressing the relative number of downstream genes
          within the corresponding sub-graph directly connected by
          directed links. This score can be interpreted as an upper
          limit of how much of the sub-graph's perturbation could have
          been introduced by the respective source, and thereby could
          be helpful to prioritize different identified sources for
          larger sub-graphs. In the same
        spirit as sources, the most downstream genes of the
        modulated sub-graphs are identified and defined as sinks. Due
        to loops not all sub-graphs are guaranteed to have sources or
        sinks. The reliability of each source is inspected using a t-test on the \textit{p}-values . The significant difference between the two groups, genes downstream the source to the genes upstream of the source is determined. This would help to assess if the source identified is reliable and marks the border between two different regulation regimes.

@@

	\texttt{MSF} generates a directed network file as an output,
        containing complete directed interactions of all modulated
        sub-graphs identified. This file could be imported into
        Cytoscape~\cite{Cyto} for visualization. Additionally, a
          file containing details on all sources and sinks for all
          modulated sub-graphs is reported. Furthermore, for
          facilitated visualization in Cytoscape, a node attribute
          file is provided, containing the source weight-age and the
          log-fold chances of all considered genes.

@@

	To demonstrate its usefulness, \texttt{MSF} is applied to
        RNA-seq data set of primary human monocyte derived macrophages
        (MDMs) infected with Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to the
        Filoviridea family: filamentous, enveloped and single stranded
        RNA viruses. EBOV causes hemorrhagic fever in humans, inducing
        the host innate and adaptive immune response to be unable to
        control virus infection~\cite{Prins}. Currently, there are no
        approved antiviral drugs for the treatment of Ebola virus
        infection~\cite{Konde,Rhein}.  The initial targets of EBOV are
        the macrophages and dendritic immune
        cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate
        immune response of the host, which includes the activation of
        alpha/beta interferon (IFN-$\alpha /
        \beta$)~\cite{Prins,Konde,Cardenas}. It has been proposed that
        IFN-$\alpha / \beta$ should be tested against Ebola for its
        antiviral activity through clinical trials~\cite{Konde}. Ebola
        infection data was selected to test the approach because it
        has been well recognized for the last several decades, and
        vast literature is available for the pathogenesis of Ebola,
        thereby, facilitating the verification of the results of
        \texttt{MSF} with the vast literature present on Ebola
        infection. Especially, the detection of IFN-$\alpha / \beta$
        as point of action for the virus, could be considered as a
        basic indicator of the correctness and usefulness of the
        approach.

@@

	Table~\ref{tab:rawVsHartung} compares the results of
        \texttt{MSF}, namely the number of detected sub-modules and
        their total genes numbers, to a simple analysis of mapping
        significantly modulated genes from the DGEA to the network and
        joining neighbors to modules. The numbers indicate that
        \texttt{MSF} detects less but larger and interpretable
        sub-modules, applying its statistical test. Furthermore, the
        dependency of the results from the \textit{p}-value cutoff
        choice is demonstrated for the DGEA, which is avoided for
        \texttt{MSF} altogether. It showcases how applying different
        cut-offs to the \textit{p}-value of genes from edgeR to the
        sub-graphs identified by \texttt{MSF} breaks the larger
        sub-graphs to many smaller unconnected sub-graphs, many of
        which are single genes.

@@

	\subsection*{Robustness}Gene set enrichment analysis

	A potential concern is how noise in the gene expression
        measurements affects our analysis. To assess the robustness
        and stability of our method, we therefore added Poisson
        distributed noise to the read counts of the three time-points
        data set, used above. Then DGEA was carried out on the
        disturbed data with the same parameters as for the native data
        using edgeR, followed by analysis with \texttt{MSF}. This
        procedure was carried out 100 times.  Every time the genes
        from the modulated sub-graphs identified from noisy data were
        compared to the genes of sub-graphs identified from the native
        data. The robustness of \texttt{MSF} analysis for the
        time-point 6, 24, and 48~hpi is shown in
        Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was
        modeled can be considered as rather stringent as \texttt{MSF}
        is sensitive to \textit{p}-value changes across the whole
        range of possible values. The observed median recall rates lay
        between 71~\% (6~hpi) and 84~\% (48~hpi).Gene set enrichment analysis

@@

	All enriched pathways with a cut-off of \textit{p}-value
        < 0.05 for \texttt{MSF} and DEG list for the
        three time-points were selected. The comparison shows most of
        the pathways known from literature to be dis-regulated by
        Ebola infection are enriched in both the enrichment
        analysis. Toll-Like receptor signaling pathway when interacts
        with EBOV glycoprotein (GP), it triggers the activation of
        cytokines~\cite{Olejnik}. Toll-like receptor pathway is
        expected to be dis-regulated in the early stage of infection,
        this pathway was not identified as significantly dis-regulated
        when \textit{p}-value cut off DEG list was analyzed for
        enrichment. Nine Toll-like receptor cascades TLR10, TLR2,
        TLR3, TLR4, TLR5, TLR7/8, TLR9, TLR1:TLR2 and TLR6:TLR2 were
        identified as dis-regulated from gene enrichment analysis of
        \texttt{MSF} identified sub-graph genes, not a single one of
        these cascade was shown to be dis-regulated in pathway
        enrichment analysis from DEG cut-off list. Since \texttt{MSF}
        considers the complete DEG results, even the weak signal at
        the earliest time-point was detected; for example Toll-like
        receptor signaling. While \texttt{MSF} is able to catch weak
        signals, it does not provide information about the functional
        relationships among genes like Reactome tool.

@@

	Gene set enrichment analysis (GSEA)~\cite{Subramanian15545}
        is a method to identify
        classes of genes or proteins that are over-represented in a
        large set of genes or proteins. GSEA uses statistical
        approaches to identify significantly enriched or depleted
        groups of genes. The complete DEG list from DGEA of the first
        time-point 6 hpi was analyzed using bioconductor package
        GSEABase (version 1.44.0). GSEA was able to identify
        Toll-like receptor, chemokine signaling pathway,
        cytosolic DNA-sensing pathway, Jak-STAT signaling pathway,
        RIG-I-like receptor signaling pathway and apoptosis as the
        highest ranked pathways. Although GSEA identified the
        important pathways for Ebola infection, it did not show how
        the cross-talk and topology of genes between the pathways
        identified. \texttt{MSF} and GSEA uses complete DEG list
        without any cut-off, is why pathways important for Ebola
        infection showed up even with weak signal genes in it.

@@

	\texttt{MSF} is a fast and easy to use tool to find
        concertedly modulated sub-graphs in a given network. Its
        implementation in java enables its use across many operating
        systems e.g. linux and windows. So far the raw output from
        edgeR~\cite{edgeR} and DESeq2~\cite{love2014moderated} are
        supported.

>>>>

		To overcome this limitation, we developed a novel
                approach to identify concertedly modulated sub-graphs
                in the global cell signaling network, based on the
                DGEA results of all genes tested. To this end,
                expression patterns of genes are integrated according
                to the topology of their interactions and allow
                potentially to read the flow of information and
                identify the effectors. The described software, named
                \texttt{Modulated Sub-graph Finder} (\texttt{MSF}) is
                freely available at
                \url{https://github.com/Modulated-Subgraph-Finder/MSF}.

@@

	Existing pathway-based analysis approaches use different
        research designs, which can be categorized into ORA
        (Over-representation analysis), FCS (Functional class scoring)
        and pathway topology based methods. All of which aim to find a
        subset of genes, e.g., significantly differentially expressed
        genes, genes associated with a certain pathway more often than
        expected given the total set of examined genes, e.g.~the whole
        genome background. ORA is the first and the most basic method
        of pathway analysis~\cite{Campos}. It uses a DEG list with
        user defined cut-off for the log-fold change and
        \textit{p}-value (most commonly using absolute log-fold change
        $\geq$ 2 and \textit{p}-value $\leq$ 0.05). Subsequently, sets
        of genes associated with annotated pathways are tested for
        being over-represented in the set of DEGs. To this end,
        hyper-geometric distribution, chi-square tests, binomial
        probability or the Fisher’s exact test are used, whereas, the
        information of the topology of genes in the pathways is
        neglected~\cite{Bayer}. Furthermore, ORA assumes that the
        biological pathways are independent of each other and ignores
        the fact that they cross-talk and
        overlap~\cite{Khatri2012,Campos}.

@@

	From another perspective, network based approaches do not
        categorize sets of genes into functional pathways, but they
        consider all interactions to be equal. Thereby, they avoid
        distinguishing arbitrarily between interactions within a
        pathway and interactions between pathways (i.e.,
        cross-talk). With this they aim to identify sub-networks that
        show modulation between two conditions or upon a
        stimuli~\cite{mitra_carvunis_ramesh_ideker_2013}. To find
        these active modules heuristics solutions like simulated
        annealing (SA), greedy methods, genetic algorithms (GA),
        network propagation and co-clustering methods are
        used~\cite{mitra_carvunis_ramesh_ideker_2013}. jActiveModules
        has been the first of this kind using simulated annealing to
        find modulated sub-networks~\cite{jActiveModules}. The benefit
        of omitting pathways is bought by reduced interpretability of
        the results due to the lack of functional labels on the
        networks.

@@

	As input, information on functional links between genes
        provided by e.g.~KEGG or Reactome and information on the
        differential expression status of single genes resulting from
        a DGEA, are required. As a result the analysis returns
        sub-graphs and their joint confidence scores, reflecting how
        the perturbation migrates through the network. Furthermore,
        the entry points of perturbation in the networks and overlap
        with conventional pathway categories are returned. To
        facilitate prioritization of the perturbation entry points, to
        each a impact score and a measure of its reliability is
        assigned. The impact score expresses the fraction of the
        sub-module being downstream of the entry point. The
        reliability is measured using a t-test on the
        \textit{p}-values of the immediate upstream and downstream
        genes. The output is prepared in a directed adjacency file,
        convenient for visualization, e.g., with
        StringApp~\cite{StringApp}, available as a Cytoscape
        plug-in~\cite{Cyto}.

	The proposed algorithm is named \texttt{Modulated Sub-graph
          Finder} (abbreviated \texttt{MSF}). \texttt{MSF} can help
        transform the information obtained from DGEA into
        comprehensible knowledge of signal transduction of genes,
        hence being a valuable complement to existing pathway based
        methods. \texttt{MSF} is freely accessible on git hub under
        the terms of the Creative Commons Attribution 4.0
        International License.

@@

	\texttt{MSF} was implemented as a Java program. It is
        developed as a novel heuristic approach to find concertedly
        modulated sub-graphs in networks of biological
        interactions. \texttt{MSF} does not use predefined gene sets
        grouped into functional units, but rather relies purely on the
        network of interacting genes. The input network consists of
        nodes corresponding to genes and edges representing
        interactions. Furthermore it utilizes comprehensive results
        from a differential gene expression analysis to discover the
        sub-graphs, or modules, which are as a whole modulated.

@@

	Lambda $\lambda$ be the weights for each gene, currently all
        genes have equal weight, i.e.~1. The combined \textit{p}-value
        $t(\rho)$ of a sub-graph will express the significance of all
        genes in the sub-graph being modulated together. The
        information from the different genes are used as, although not
        independent, replicated measurement of the behavior of the
        whole sub-graph. This potentially increases the power to
        detect also significant sub-graphs consisting of genes which
        are not significant on there own.

@@

	In a last post processing step \texttt{MSF} identifies the
        trigger points of the modulated sub-graphs. These trigger
        genes are the sources of the sub-graphs with only outgoing
        edges. These genes can be interpreted as the possible entry
        points of perturbation from where the stimulus causes
        downstream effects. Each individual source is given an impact
        score, expressing the relative number of downstream genes
        within the corresponding sub-graph directly connected by
        directed links. This score can be interpreted as an upper
        limit of how much of the sub-graph's perturbation could have
        been introduced by the respective source, and thereby could be
        helpful to prioritize different identified sources for larger
        sub-graphs. In the same spirit as sources, the most downstream
        genes of the modulated sub-graphs are identified and defined
        as sinks. Due to loops not all sub-graphs are guaranteed to
        have sources or sinks. The reliability of each source is
        inspected using a t-test on the \textit{p}-values . The
        significant difference between the two groups, genes
        downstream the source to the genes upstream of the source is
        determined. This would help to assess if the source identified
        is reliable and indeed marks the border between two different
        regulation regimes.

@@

	\texttt{MSF} generates a directed network file as an output,
        containing complete directed interactions of all modulated
        sub-graphs identified. This file could be imported into
        Cytoscape~\cite{Cyto} for visualization. Additionally, a file
        containing details on all sources and sinks for all modulated
        sub-graphs is reported. Furthermore, for facilitated
        visualization in Cytoscape, a node attribute file is provided,
        containing the source weightage and the log-fold chances of
        all considered genes.

@@

	To demonstrate its usefulness, \texttt{MSF} is applied to
        RNA-seq data set of primary human monocyte derived macrophages
        (MDMs) infected with Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to the
        Filoviridea family: filamentous, enveloped and single stranded
        RNA viruses. EBOV causes hemorrhagic fever in humans, inducing
        the host innate and adaptive immune response to be unable to
        control virus infection~\cite{Prins}. Currently, there are no
        approved antiviral drugs for the treatment of Ebola virus
        infection~\cite{Konde,Rhein}.  The initial targets of EBOV are
        the macrophages and dendritic immune
        cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate
        immune response of the host, which includes the activation of
        alpha/beta interferon (IFN-$\alpha /
        \beta$)~\cite{Prins,Konde,Cardenas}. It has been proposed that
        IFN-$\alpha / \beta$ should be tested against Ebola for its
        antiviral activity through clinical trials~\cite{Konde}. Ebola
        infection data was selected to test the approach because it
        has been well recognized for the last several decades, and
        vast literature is available for the pathogenesis of Ebola,
        hence facilitating the verification of the results of
        \texttt{MSF} with the vast literature present on Ebola
        infection. Especially, the detection of IFN-$\alpha / \beta$
        as point of action for the virus, could be considered as a
        basic indicator of the correctness and usefulness of the
        approach.

@@

	Table~\ref{tab:rawVsHartung} compares the results of
        \texttt{MSF}, namely the number of detected sub-modules and
        their total genes numbers, to a simple analysis of mapping
        significantly modulated genes from the DGEA to the network and
        joining neighbors to modules. The numbers indicate that
        \texttt{MSF} detects less but larger and easier interpretable
        sub-modules, applying its statistical test. Furthermore, the
        dependency of the results from the \textit{p}-value cutoff
        choice is demonstrated for the DGEA, which is avoided for
        \texttt{MSF} altogether. It showcases how applying different
        cut-offs to the \textit{p}-value of genes from edgeR to the
        sub-graphs identified by \texttt{MSF} breaks the larger
        sub-graphs to many smaller unconnected sub-graphs, many of
        which are single genes.

@@

	\subsection*{Robustness}\sout{Gene set enrichment analysis}

	A potential concern is how noise in the gene expression
        measurements affects our analysis. To assess the robustness
        and stability of our method, we therefore added Poisson
        distributed noise to the read counts of the three time-points
        data set, used above. Then DGEA was carried out on the
        disturbed data with the same parameters as for the native data
        using edgeR, followed by analysis with \texttt{MSF}. This
        procedure was carried out 100 times.  Every time the genes
        from the modulated sub-graphs identified from noisy data were
        compared to the genes of sub-graphs identified from the native
        data. The robustness of \texttt{MSF} analysis for the
        time-point 6, 24, and 48~hpi is shown in
        Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was
        modeled can be considered as rather stringent as \texttt{MSF}
        is sensitive to \textit{p}-value changes across the whole
        range of possible values. The observed median recall rates lay
        between 71~\% (6~hpi) and 84~\% (48~hpi).\sout{Gene set enrichment analysis}

@@

	All enriched pathways with a cut-off of \textit{p}-value <
        0.05 for \texttt{MSF} and DEG list for the three time-points
        were selected. The comparison shows most of the pathways known
        from literature to be dis-regulated by Ebola infection are
        enriched in both the enrichment analysis. Toll-like receptor
        signaling pathway when interacts with EBOV glycoprotein (GP)\TODO{???},
        it triggers the activation of
        cytokines~\cite{Olejnik}. Toll-like receptor pathway is
        expected to be dis-regulated in the early stage of infection,
        this pathway was not identified as significantly dis-regulated
        when \textit{p}-value cut off DEG list was analyzed for
        enrichment. Nine Toll-like receptor cascades TLR10, TLR2,
        TLR3, TLR4, TLR5, TLR7/8, TLR9, TLR1:TLR2 and TLR6:TLR2 were
        identified as dis-regulated from gene enrichment analysis of
        \texttt{MSF} identified sub-graph genes, not a single one of
        these cascade was shown to be dis-regulated in pathway
        enrichment analysis from DEG cut-off list. Since \texttt{MSF}
        considers the complete DEG results, even the weak signal at
        the earliest time-point was detected; for example Toll-like
        receptor signaling. While \texttt{MSF} is able to catch weak
        signals, it does not provide information about the functional
        relationships among genes like Reactome tool.

@@

	Gene set enrichment analysis (GSEA)~\cite{Subramanian15545}
        is a method to identify
        classes of genes or proteins that are over-represented in a
        large set of genes or proteins. GSEA uses statistical
        approaches to identify significantly enriched or depleted
        groups of genes. The complete DEG list from DGEA of the first
        time-point 6 hpi was analyzed using bioconductor package
        GSEABase (version 1.44.0). GSEA was able to identify
        Toll-like receptor, chemokine signaling pathway,
        cytosolic DNA-sensing pathway, Jak-STAT signaling pathway,
        RIG-I-like receptor signaling pathway and apoptosis as the
        highest ranked pathways. Although GSEA identified the
        important pathways for Ebola infection, it did not show how \TODO{???}
        the cross-talk and topology of genes between the pathways
        identified. \texttt{MSF} and GSEA uses complete DEG list
        without any cut-off, \TODO{????} is why pathways important for Ebola
        infection showed up even with weak signal genes in it.

@@

	\texttt{MSF} is a fast and easy to use tool to find
        concertedly modulated sub-graphs in a given network. Its
        implementation in Java enables its use across many operating
        systems e.g. linux and windows. So far the raw output from
        edgeR~\cite{edgeR} and DESeq2~\cite{love2014moderated} are
        supported.
